Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-19 manuscript.tex
Section figure 5 done!

\subsection*{
  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
}
\textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells). \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each region (y-axis) and pattern (x-axis). \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grant2011fimo,schreiber2025tangermeme} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.

@@

@@

\subsection*{Figure 5. }

>>>>

\subsection*{
  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
}
\textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells). \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grant2011fimo,schreiber2025tangermeme} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.

@@

\subsection*{
  Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
}

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d}). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d}). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiran2024synthetic}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}. The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothstein2020heterodmerization}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig. 5d}).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kim2024dnaguided}.\par

Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
method we found three sets of regions that are predicted to be contain instances of different TFs. One set where
instances of TEAD4, GRHL1/2 and ZEB1/2 co-occur, one where instances of Nuclear receptor, TFAP2A/B/C and ZIC1-5
instances co-occur and one set where instances of FOXC2 and TWIST1/ALX4 co-occur.\par

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{white2021insights}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{white2021insights} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models. Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G>A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp upsteam of a partial homeobox site potentially creating a new binding site for TWIST1-ALX4 dimers (\textbf{Fig. 5h}).\par

@@

\subsection*{
  Figure 5. Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
}
\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics. \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{white2021insights} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome. Lead SNPs are shown with a bigger dot size. \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red. \textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube_{organoid}$ of the references (top) and alternative (bottom) allele of SNP rs1555067. The location of the SNP is indicated with a dotted line.

\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_6.png}

\subsection*{Figure 6. }

\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_7.png}

\subsection*{Figure 7. }
